Experience Report: Haskell in Computational Biology

### 2012-03-11 mrfy_experience_report.tex
edites through section 3.1

@@

\newcommand\txprobj[3][]{a#1_{#2_{j-1}#3_j}}
\newcommand\txprobjj[3][]{a#1_{#2_{j-1}#3_j}}
\newcommand\alignwidth{\ensuremath R} % number of columns in an alignment
\newcommand\pairedwith[1]{{\pi(#1)}}

@@

Homology-detection software is most often used in one of two ways:
to test a hypothesis about 
the function of a single, newly discovered protein, or 
to compare every protein in a genome against a library of known protein 
structures.
Either way, 
the software is \emph{trained}
on a group of proteins that share function and structure.
These proteins are identified by a biologist, who puts
their amino-acid sequences into a correspondence, also called an
\emph{alignment}. 
This alignment represents structural
correspondence among individual amino acids in a set of homologous proteins.
An~alignment may be represented as a matrix
(Figure~\ref{alignment}) 
in which each row corresponds to the amino-acid sequence of a protein,
and each column groups amino acids that play similar roles in
different proteins.

An alignment may contain \emph{gaps} (notated as dashes in
\figref{alignment}) for some proteins.
A~gap in row~2, column~$j$ indicates that as proteins evolved, either 
protein~2 lost its amino acid in position~$j$, or 
other proteins gained an amino acid in position~$j$.
If~column~$j$ contains few gaps, 
it~is considered a \emph{consensus column},
and the few proteins with gaps probably lost amino acids via
\emph{deletions}.\remark{NMD: Missing is the fact that all these models
are directionless; they don't care whether something was gained or lost
over time. Perhaps this doesn't matter.}
If~column~$j$ contains \emph{mostly} gaps, 
it~is considered a \emph{non-consensus column},
and the few proteins without gaps probably gained amino acids via
\emph{insertions}.

@@

\caption{A~structural alignment of four proteins $(\alignwidth = 4)$}
\label{alignment} 
\end{figure}

A~hidden Markov model carries probabilities on its states and on its
state transitions.
Both the probabilities and the
the states are determined by the alignment:
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}~$M_j$.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has amino acid~$x$ in
 column~$j$.
This probability is called an \emph{emission} probability.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}~$D_j$.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}~$I_j$.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained amino acid~$x$ by
insertion at column~$j$.
\end{itemize}
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
We~represent a state using an algebraic data type:
\begin{smallverbatim}
data HMMState = Mat | Ins | Del | Beg | End
\end{smallverbatim}

The particular hidden Markov model we use requires that every nonempty
sequence of insertions be followed by a match, and similarly for
deletions.
\remark{citation needed?}
In~other words, insertions and deletions cannot be mixed without an
intervening match.
This restriction means that between two adjacent columns, six
transitions are possible.
An~insertion state may also transition to itself.
\begin{itemize} 
\item
A~transition into a match state in
column~$j$ is more likely when column~$j$ is a consensus column.
Depending on the predecessor state, 
the probability of such a transition is notated
$\txprobj M M$, $\txprobj I M$, or $\txprobj D M$.
\item
A~transition into a deletion state in
 column~$j$ is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is notated
$\txprobj M D$ or $\txprobj D D$.
\item
A~transition into an insertion state in
 column~$j$ is relatively unlikely,
but it is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is notated
$\txprobjj M I$ or $\txprobjj I I$.
\end{itemize}

@@

Given a hidden Markov model, 
an~established software package called HMMER (``hammer'') 
can compute the probability
that a new protein shares structure and
evolutionary history with the proteins used to train the model.
The computation finds the most likely path through the hidden Markov model.
To~make best use of floating-point arithmetic, the software computes
the \emph{logarithm} of the probability of each path, by summing
the logs of the 
probabilities on the states and edges of the path \cite{Viterbi:1967hq}.
(The logarithm function is monotonic, so the path that maximizes the
log of the probability is the most likely path.)
The~specification is shown on the left-hand side of \figref{viterbi}.
The probability $V_j^M(i)$ represents the probability of the most
likely path of the first~$i$ amino acids in the query sequence,
terminating with placement of amino acid $x_i$ in state~$M_j$.
The probabilities $V_j^D(i)$ and $V_j^D(i)$ are similar.
Equations like these can be found in more recent literature
\cite{Durbin:1998wz,Eddy:1998ut}.

To~be able to use Haskell, we had to reimplement the standard
algorithm for solving Viterbi's equations.
Haskell made it possible for us to write code that looks like the
math,
which made the code easy to write and gives us confidence that it is
correct.
We~begin with our representations.

In~idiomatic Haskell, 
we might represent an individual amino acid~$x$
using a value of algebraic data type, thus:
\begin{smallverbatim}
data Amino = Ala | Cys | Asp | Glu | Gly | ...
\end{smallverbatim}
But our primary goal was to get something working quickly with a
legacy file format that represents each amino acid as a small integer.
Our representation is therefore
\begin{smallverbatim}
type AA       = Int -- amino acid using HMMER coding
\end{smallverbatim}
Either way, a query sequence is (and should be) an immutable vector of
amino acids.
%type QuerySeq = Vector AA
%\end{smallverbatim}

The log of a probability is simply a number.
To distinguish a transition probability from other numbers, we wrap it in
the value constructor \texttt{TProb}:
\begin{smallverbatim}
type LogProb = Double
data TProb   = TProb { logProbability :: LogProb }
\end{smallverbatim}
We group transition probabilities into 7-tuples.
Each tuple corresponds to a ``node'' in the hidden Markov model
(a~dashed ovel in \figref{plan7}), and to a column of the original
alignment.
As~noted above, 
a match state may transition to any state in the successor node;
an insertion state may transition to itself or to its successor match
state;
and
a~deletion may transition to its successor match
state or deletion state.
\begin{smallverbatim}
data TProbs = 
     TProbs { m_m :: TProb, m_i :: TProb, m_d :: TProb
            , i_m :: TProb, i_i :: TProb
            , d_m :: TProb, d_d :: TProb }
\end{smallverbatim}

The representation of a node includes the probabilities of transitions
into the three states of that node, plus the tables
of emission probabilities for the match and insertion states.
\begin{smallverbatim}
type EProbs  = Vector LogProb
data HmmNode = HmmNode { transitions  :: TProbs
                       , matEmissions :: EProbs
                       , insEmissions :: EProbs
                       }
\end{smallverbatim}

In~Viterbi's equations,
the probability in each state is a simple function of the probability
in its predecessor states, 
and all probabilities can be computed by a classic dynamic-programming
algorithm
which starts at the begin state,
computes probabilites in nodes $1$~through~$\alignwidth$ in
succession, and terminates at the end state.
One of us implemented this algorithm, storing the probabilities in an array.
\remark{Why is the number of dimensions 3 and not 2 (indices $i$, $j$,
and what??) and why do we care?}
The cost was
$O(|N|\times|\alignwidth|)$;
in our work, $\alignwidth$~and~$N$ range from several hundred to a few
thousand.

@@

Another of us was curious to try to code Viterbi's equations
directly as recursive functions.
Like the classic recursive Fibonacci function, Viterbi's functions,
when implemented \naive ly,
perform extraordinarily badly.
But just like the Fibonacci function, Viterbi's functions can be
\emph{memoized}.
To~show the memoized code, we have to confess to a small
transformation of Viterbi's equations.
The logarithms on the left of \figref{viterbi} are computed when the
hidden Markov model is constructed, and the resulting numbers are
stored in a file.
Moreover, to make the file more pleasant to read, the logarithms are
negated.
(The logarithm of a probability is typically negative, and negating
the logarithm eliminates hordes of minus signs from the file.)
We~therefore implement a transformed version of Viterbi's equations,
shown on the right-hand side of \figref{viterbi}, which minimize the
negated log probability for each combination of node~$j$, amino
acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.

Our implementation computes not only the minimum
negated log probability, called a \texttt{Score},
but also the path of states which leads to that most likely outcome. 
Elsewhere, MRFy pairs not only paths but also other types of values
with scores, so we have defined a small abstraction:
\smallverbatiminput{vscore}
We~made \texttt{Scored} an instance of~\texttt{Ord} and of
\texttt{Functor}.
Think of a value of type \texttt{Scored~a} as a container holding
 an~``\texttt a'' with a number written on the side.
The \texttt{/+/} function adds to the number without touching the
 container.
Function 
\begin{smallverbatim}
fmap :: (a -> b) -> (Scored a -> Scored b)
\end{smallverbatim}
applies a function to what's inside the container, without touching
the number.
And containers are ordered purely by number, so applying
\texttt{minimum} to a list of scored things chooses the thing with the
smallest (and therefore best) score.

Given the \texttt{Scored} abstraction, we can easily implement
Viterbi's equations for a match state.
We~choose the best solution
from among the three predecessor states;
we~use~\texttt{/+/} to add transition and emission scores,
and we use \texttt{fmap} to add \texttt{Mat} to the list of states
encountered up to this point:
\remark{TODO: (1)~Call the inner function \texttt{vee}.
(2)~Use \texttt{minimum (map from [Mat, Ins, Del])}
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\smallverbatiminput{vfix}
Functions \texttt{tProb} and \texttt{edge} combine to select the
appropriate transition score (one of the $a'$~values).
% Similarly, the insertion equation is:
% \begin{verbatim}                    
%   -- consume an observation but not a node
%   | s == ins = let trans t prev =
%                    (score + tProb n t + (eProb n o), ins:path)
%                     where (score, path) = viterbi prev n (o - 1)
%                 in minimum [
%                     trans m_i mat,
%                     trans i_i ins
%                     ]
% \end{verbatim}
% Finally, the deletion equation is:
% \begin{verbatim}                    
%   -- consume a node but not an observation
%   | s == del = let trans t prev = 
%                       (score + tProb (n-1) t, del:path)
%                       where (score, path) = viterbi prev (n - 1) o
%                 in minimum [
%                     trans m_d mat,
%                     trans d_d del
%                     ]
% \end{verbatim}
Programming the algorithm in this form was not only intellectually
satisfying; keeping the code close to the math also helped us find
bugs.
The most memorable bug was that in one of our base cases, 
we~mistakenly returned an empty path of states;
the correct result was a singleton path containing the initial state.
Very quickly after observing a missing state in the output, we were
able to find the faulty case in the code.

Whatever the attractions of the code above, in our work we also
require high performance.
To make the simple recursive code efficient, we used Luke Palmer's
\texttt{Data.Memocombinators} package.
The function \verb+viterbi'+ is unmemoized,
and it makes recursive calls to \verb+viterbi+, which is memoized:
\begin{smallverbatim}
viterbi state j i = 
  Memo.memo3 (Memo.arrayRange (Mat, End)) 
             (Memo.arrayRange (0, numNodes))
             (Memo.arrayRange (0, seqlen)) 
             viterbi' state j i
\end{smallverbatim}
The value
\texttt{numNodes} represents~$\alignwidth$,
and \texttt{seqlen} represents~$N$.\remark{I hope this is right. ---NR}

This memoized recursive code performs just as well as the classic
dynamic-programming code.
And~the short fragment above is the \emph{only} part of the code
devoted to improving performance.
By contrast, standard implementations of the Viterbi algorithm, such as in HMMER,
spend much of their code 
managing dynamic-programming tables.
Haskell enabled us write simple, performant code with little effort.
%
Because the memoized version so faithfully resembles the equations in
\figref{viterbi}, we~retired the classic dynamic-programming version.

\remark{NR doesn't think the bug story is that interesting.}

>>>>

\newif\ifpagetuning

\pagetuningtrue % trying to get page breaks in the best spots

@@

\newcommand\txprobj[3][]{a#1_{#2_{j-1}#3_j}}
\newcommand\txprobjj[3][]{a#1_{#2_{j-1}#3_j}}
\newcommand\alignwidth{\ensuremath C} % number of columns in an alignment
\newcommand\pairedwith[1]{{\pi(#1)}}

\ifpagetuning
  \newcommand\tuningbox{\vbox}
\else
  \newcomming\tuningbox{\relax}
\fi

@@

Homology-detection software is most often used in one of two ways:
to test a hypothesis about 
the function of a single, newly discovered protein, or 
to compare every protein in a genome against a library of known protein 
structures.
Either way, 
the software is \emph{trained}
on a group of proteins that share function and structure.
These proteins are identified by a biologist, who puts
their amino-acid sequences into a correspondence, also called an
\emph{alignment}. 
This alignment relates individual amino acids in a set of homologous proteins.
An~alignment may be represented as a matrix
in which each row corresponds to the amino-acid sequence of a protein,
and each column groups amino acids that play similar roles in
different proteins (Figure~\ref{alignment}).

An alignment may contain \emph{gaps}, which in 
\figref{alignment} are shown as dashes.
A~gap in row~2, column~$j$ indicates that as proteins evolved, either 
protein~2 lost its amino acid in position~$j$, or 
other proteins gained an amino acid in position~$j$.
If~column~$j$ contains few gaps, 
it~is considered a \emph{consensus column},
and the few proteins with gaps probably lost amino acids via
\emph{deletions}.
%%  \remark{NMD: Missing is the fact that all these models
%%  are directionless; they don't care whether something was gained or lost
%%  over time. But this doesn't matter.}
If~column~$j$ contains \emph{mostly} gaps, 
it~is considered a \emph{non-consensus column},
and the few proteins without gaps probably gained amino acids via
\emph{insertions}.

@@

\caption{A~structural alignment of four proteins $(\alignwidth = 12)$}
\label{alignment} 
\end{figure}

A~hidden Markov model carries probabilities on some states and on all
state transitions.
Both the probabilities and the
the states are determined by the alignment:
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}~$M_j$.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has amino acid~$x$ in
 column~$j$.
This probability is called an \emph{emission} probability.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}~$D_j$.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}~$I_j$.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained amino acid~$x$ by
insertion at column~$j$.
\end{itemize}
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
We~represent a state using an algebraic data type:
\begin{smallverbatim}
data HMMState = Mat | Ins | Del | Beg | End
\end{smallverbatim}

The particular hidden Markov model we use requires that every nonempty
sequence of insertions be followed by a match, and similarly for
deletions.
In~other words, insertions and deletions cannot be mixed without an
intervening match.
This restriction means that there are only seven possible transitions
into the states of any one column \cite{Eddy:1998ut}. 
\begin{itemize} 
\item
A~transition into a match state in
column~$j$ is more likely when column~$j$ is a consensus column.
Depending on the predecessor state, 
the probability of such a transition is 
$\txprobj M M$, $\txprobj I M$, or $\txprobj D M$.
\item
A~transition into a deletion state in
 column~$j$ is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobj M D$ or $\txprobj D D$.
\item
A~transition into an insertion state in
 column~$j$ is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobjj M I$ or $\txprobjj I I$.
\end{itemize}

@@

Given a hidden Markov model, 
an~established software package called HMMER (pronounced ``hammer'') 
can compute the probability
that a new protein shares structure and
evolutionary history with the proteins used to train the model.
The computation finds the most likely path through the hidden Markov model.
To~make best use of floating-point arithmetic, the software computes
the \emph{logarithm} of the probability of each path, by summing
the logs of the 
probabilities on the states and edges of the path \cite{Viterbi:1967hq}.
The logarithm function is monotonic, so the path that maximizes the
log of the probability is the most likely path.

The computation is specified on the left-hand side of \figref{viterbi}.
A~probability $V_j^M(i)$ represents the probability of the most
likely path of the first~$i$ amino acids in the query sequence,
terminating with placement of amino acid $x_i$ in state~$M_j$.
Probabilities $V_j^D(i)$ and $V_j^D(i)$ are similar.
These
equations are very clearly explained by
\citet[Chapter~5]{Durbin:1998wz}

To~be able to use Haskell, we had to reimplement the standard
algorithm for solving Viterbi's equations.
Haskell made it possible for us to write code that looks like the
math,
which made the code easy to write and gives us confidence that it is
correct.
To~present some code,
we~begin with representations.

In~idiomatic Haskell, 
we might represent an individual amino acid~$x$
using a value of algebraic data type:
\begin{smallverbatim}
data Amino = Ala | Cys | Asp | Glu | Gly | ...
\end{smallverbatim}
But our primary goal was to get something working quickly,
and our hidden Markov models are stored using a
legacy file format that represents each amino acid as a small integer.
Our representation is therefore
\begin{smallverbatim}
type AA = Int -- amino acid using HMMER coding
\end{smallverbatim}
Either way, a query sequence is (and should be) an immutable vector of
amino acids.
%type QuerySeq = Vector AA
%\end{smallverbatim}

The log of a probability is just a number.
To distinguish a transition probability from other numbers, we wrap it in
a value constructor:
\begin{smallverbatim}
type LogProb = Double
data TProb   = TProb { logProbability :: LogProb }
\end{smallverbatim}
We group transition probabilities into 7-tuples.
Each tuple corresponds to a ``node'' in the hidden Markov model
(a~dashed oval in \figref{plan7}) and to a column of the original
alignment.
As~noted above, 
a match state may transition to any state in the successor node;
an insertion state may transition to itself or to its successor match
state;
and
a~deletion may transition to its successor match
state or deletion state.
\begin{smallverbatim}
data TProbs = 
     TProbs { m_m :: TProb, m_i :: TProb, m_d :: TProb
            , i_m :: TProb, i_i :: TProb
            , d_m :: TProb, d_d :: TProb }
\end{smallverbatim}
In~our code, we select from this 7-tuple using the \texttt{edge}
function:
\smallverbatiminput{edge}
In MRFy, the first two arguments to \texttt{edge} are known at compile
time, and we expect \texttt{edge} to be inlined.

The representation of a node includes the probabilities of transitions
into the states of that node, plus the tables
of emission probabilities for the match and insertion states.
\begin{smallverbatim}
type EProbs  = Vector LogProb
data HmmNode = HmmNode { transitions  :: TProbs
                       , matEmissions :: EProbs
                       , insEmissions :: EProbs
                       }
\end{smallverbatim}

In~Viterbi's equations,
the probability in each state is a function of the probabilities
in its predecessor states, 
and all probabilities can be computed by a classic dynamic-programming
algorithm.
This algorithm starts at the begin state,
computes probabilites in nodes $1$~through~$\alignwidth$ in
succession, and terminates at the end state.
One of us implemented this algorithm, storing the probabilities in an array.
The cost was
$O(|N|\times|\alignwidth|)$;
in~MRFy, $\alignwidth$~and~$N$ range from several hundred to a few
thousand.

@@

Another of us was curious to try coding Viterbi's equations
directly as recursive functions.
Like the classic recursive Fibonacci function, Viterbi's functions,
when implemented \naive ly,
perform extraordinarily badly.
But just like the Fibonacci function, Viterbi's functions can be
\emph{memoized}.
To~show our code, we have to confess to a 
transformation of Viterbi's equations.
The~logarithms on the left of \figref{viterbi} are computed when the
hidden Markov model is constructed, and the resulting numbers are
stored in a file.
And~to~make this file more pleasant to read, the logarithms are
negated.
(The logarithm of a probability is typically negative, and negating
the logarithm eliminates hordes of minus signs.)
We~therefore implement a transformed version of Viterbi's equations,
shown on the right-hand side of \figref{viterbi}.
The transformed equations minimize the
negated log probability for each combination of node~$j$, amino
acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.
A~negated log probability is called a~\emph{score}.

MRFy computes not only the minimum score but
also the path of states which produces that score.
Scores can be usefully paired with many types of values,
so we have defined a small abstraction:
\smallverbatiminput{vscore}
Think of a value of type \texttt{Scored~a} as a container holding
 an~``\texttt a'' with a score written on the side.
The \texttt{/+/} function adds to the score without touching the
 container.

\tuningbox{
We~also made \texttt{Scored} an instance of~\texttt{Ord} and of
\texttt{Functor}, so
\begin{smallverbatim}
fmap :: (a -> b) -> (Scored a -> Scored b)
\end{smallverbatim}
applies a function to what's inside the container, without touching
the score.
And containers are ordered by score alone, so applying
\texttt{minimum} to a list of scored things chooses the thing with the
smallest (and therefore best) score.
}

Given the \texttt{Scored} abstraction, we can easily implement
Viterbi's equations for a match state.
To~implement the equation at the top right of \figref{viterbi},
we~choose the best solution
from among the three predecessor states;
we~use~\texttt{/+/} to add transition and emission scores,
and we use \texttt{fmap} to add \texttt{Mat} to the list of states
encountered up to this point:
\remark{TODO: 
(1)~Call the inner function \texttt{vee}
(2)~Use \texttt{minimum (map from [Mat, Ins, Del])}
(3)~Explain the recursive call
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\smallverbatiminput{vfix}
Functions \texttt{tProb} and \texttt{edge} combine to select the
appropriate transition score (one of the $a'$~values).
Implementing the equation in this form was not only intellectually
satisfying; it~also helped us find
bugs.\remark{NR doesn't think the bug story is that interesting.}
The most memorable bug was that in one of our base cases, 
we~mistakenly returned an empty path of states;
the correct result was a singleton path containing the initial state.
Very quickly after observing a missing state in the output, we were
able to find the faulty case in the code.

Whatever the attractions of the code above, our work also
requires high performance.
To make the simple recursive code efficient, we used Luke Palmer's
\texttt{Data.Memocombinators} package.
The function \verb+viterbi'+ is unmemoized,
and it makes recursive calls to \verb+viterbi+, which is memoized:
\smallverbatiminput{memo}
The value
\texttt{numNodes} represents~$\alignwidth$,
and \texttt{seqlen} represents~$N$.\remark{I hope this is right. ---NR}

This memoized recursive code performs just as well as the classic
dynamic-programming code.
And~the call to \texttt{Memo.memo3} is the \emph{only} part of the code
devoted to improving performance.
By~contrast, standard implementations of the Viterbi algorithm, such as in HMMER,
spend much of their code 
managing dynamic-programming tables.
Haskell enabled us write simple, performant code with little effort.
%
Because the memoized version so faithfully resembles the equations in
\figref{viterbi}, we~retired the classic dynamic-programming version.
